Lessons learned: creating tutorials teaching agent-based modelling for archaeologists

### 2025-03-31 paper.Rmd
update paper based on review

While agent-based modelling importance in archaeological practice is steadily growing [@romanowska2023] there are significant barriers to its wider adoption. One of the problems is the access to vocational training required to gain the necessary computer programming and modelling skills [@daems2020]. A survey of practitioners [@davies2018] found significant deficiencies in the provision of ABM training; 76% of modellers were primarily self-taught with some degree of secondary support, i.e. 27% relied on peer support, only 7% had any sort of formal training at a workshop or summer school, while only 14% received training as part of a degree.

@@

*From review: Although the point can be somewhat interpreted throughout the manuscript, I would strongly suggest that the authors make a brief discussion of the experience (i.e. the iterative design and testing of educational material),*

@@

*From review: particularly in light of shortcomings and resistances to digital (and computational) skills in archaeology.*

>>>>

While agent-based modelling importance in archaeological practice is steadily growing [@romanowska2023] there are significant barriers to its wider adoption. While education on things like GIS and databases have found their way into many study programs [e.g. @visser2016; @garstki2022], the access to vocational training required to gain the necessary computer programming and modelling skills is limited. A survey of practitioners [@davies2018] found significant deficiencies in the provision of ABM training; 76% of modellers were primarily self-taught with some degree of secondary support, i.e. 27% relied on peer support, only 7% had any sort of formal training at a workshop or summer school, while only 14% received training as part of a degree. While ABM is taught outside of archaeology [@augustijn2019], it remains lacking in most curricula due to for example lack of expertise in a university or pressure on the curriculum due to other subjects to be taught. The lack of training material on ABM for archaeologists, leaves this important method in digital archaeology at the "fringes of the archaeological discipline" [@daems2020, 90] rather than moving to the front lines.

@@

The chosen workflow by developing tutorials. testing these ourselves and then testing these during workshops and conferences has proven very valuable in many ways. First of all it had a very practical use. New users found bugs that we as a team missed or did not encounter. Participants used different operating systems, different web browsers and different settings on their computers, leading to other uses and other bugs. In addition, different people do different things on their computers. This behavioral differences between users helped us to understand how people would use our tutorials and enabled improvements. Thirdly, our presence on conferences and organizing these workshops improved dissemination of the tutorials. A fourth aspect is related, since the high number of people interested in these workshops has shown the need and demand for educational material for ABM.

@@

Based on the high demand for our online and offline workshops, it might be possible to conclude that "resistance to digital tools" [@morgan2022, 216] might be dwindling. The demand for our tutorials and workshops, shows that there is an increasing group of archaeologists learning to learn a new tool. At the same time the tutorials help to move ABM from the fringes of archaeology in offering a easy access to ABM and programming. However, we also noted that some participants found the workshops interesting, but would not pursue to learn ABM any further. However, these will probably never apply ABM in their research, although they are now aware of the possibilities of the method, at least acknowledging the possibilities of ABM and digital archaeology in de broader sense.

Teaching digital archaeology is often related to the the staff of an institute, although some institutes are incorporating digital aspects of archaeology in the regular curriculum [e.g. @visser2016; @lambers2024]. The number of specialists in ABM is limited world wide and the OER's created in this project enable archaeologists an easy introduction into ABM, modelling and computational thinking. However, we encourage people to discuss their ideas with other people in the field, for example during CAA conferences (<https://caa-international.org/>). We have seen during the workshops that discussion is probably the most important aspect of model building, which is impossible to incorporate in self-paced learning.

### 2025-03-31 paper.tex
update paper based on review

\title{Lessons learned: creating tutorials teaching agent-based modelling for archaeologists}
\author{}
\date{\vspace{-2.5em}March 17, 2025}

@@

While agent-based modelling importance in archaeological practice is steadily growing (Romanowska and Scherjon 2023) there are significant barriers to its wider adoption. One of the problems is the access to vocational training required to gain the necessary computer programming and modelling skills (Daems 2020). A survey of practitioners (Davies and Romanowska 2018) found significant deficiencies in the provision of ABM training; 76\% of modellers were primarily self-taught with some degree of secondary support, i.e.~27\% relied on peer support, only 7\% had any sort of formal training at a workshop or summer school, while only 14\% received training as part of a degree.

@@

\emph{From review: Although the point can be somewhat interpreted throughout the manuscript, I would strongly suggest that the authors make a brief discussion of the experience (i.e.~the iterative design and testing of educational material),}

@@

\emph{From review: particularly in light of shortcomings and resistances to digital (and computational) skills in archaeology.}

@@

@@

@@

@@

@@

>>>>

\title{Lessons learned: creating tutorials teaching agent-based modelling for archaeologists}
\author{}
\date{\vspace{-2.5em}March 31, 2025}

@@

While agent-based modelling importance in archaeological practice is steadily growing (Romanowska and Scherjon 2023) there are significant barriers to its wider adoption. While education on things like GIS and databases have found their way into many study programs (Garstki 2022; e.g. Visser, Zijverden and Alders 2016), the access to vocational training required to gain the necessary computer programming and modelling skills is limited. A survey of practitioners (Davies and Romanowska 2018) found significant deficiencies in the provision of ABM training; 76\% of modellers were primarily self-taught with some degree of secondary support, i.e.~27\% relied on peer support, only 7\% had any sort of formal training at a workshop or summer school, while only 14\% received training as part of a degree. While ABM is taught outside of archaeology (Augustijn et al. 2019), it remains lacking in most curricula due to for example lack of expertise in a university or pressure on the curriculum due to other subjects to be taught. The lack of training material on ABM for archaeologists, leaves this important method in digital archaeology at the ``fringes of the archaeological discipline'' (Daems 2020: 90) rather than moving to the front lines.

@@

The chosen workflow by developing tutorials. testing these ourselves and then testing these during workshops and conferences has proven very valuable in many ways. First of all it had a very practical use. New users found bugs that we as a team missed or did not encounter. Participants used different operating systems, different web browsers and different settings on their computers, leading to other uses and other bugs. In addition, different people do different things on their computers. This behavioral differences between users helped us to understand how people would use our tutorials and enabled improvements. Thirdly, our presence on conferences and organizing these workshops improved dissemination of the tutorials. A fourth aspect is related, since the high number of people interested in these workshops has shown the need and demand for educational material for ABM.

@@

Based on the high demand for our online and offline workshops, it might be possible to conclude that ``resistance to digital tools'' (Morgan 2022: 216) might be dwindling. The demand for our tutorials and workshops, shows that there is an increasing group of archaeologists learning to learn a new tool. At the same time the tutorials help to move ABM from the fringes of archaeology in offering a easy access to ABM and programming. However, we also noted that some participants found the workshops interesting, but would not pursue to learn ABM any further. However, these will probably never apply ABM in their research, although they are now aware of the possibilities of the method, at least acknowledging the possibilities of ABM and digital archaeology in de broader sense.

Teaching digital archaeology is often related to the the staff of an institute, although some institutes are incorporating digital aspects of archaeology in the regular curriculum (Lambers 2024; e.g. Visser, Zijverden and Alders 2016). The number of specialists in ABM is limited world wide and the OER's created in this project enable archaeologists an easy introduction into ABM, modelling and computational thinking. However, we encourage people to discuss their ideas with other people in the field, for example during CAA conferences (\url{https://caa-international.org/}). We have seen during the workshops that discussion is probably the most important aspect of model building, which is impossible to incorporate in self-paced learning.

@@

\leavevmode\vadjust pre{\hypertarget{ref-augustijn2019}{}}%
Augustijn, E-W, Kounadi, O, Kuznecova, T and Zurita-Milla, R. 2019 \emph{Teaching Agent-Based Modelling and Machine Learning in an integrated way. GeoComputation 2019} DOI: https://doi.org/\href{https://doi.org/10.17608/k6.auckland.9848804}{10.17608/k6.auckland.9848804}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-garstki2022}{}}%
Garstki, K. 2022 Teaching for Data Reuse and Working toward Digital Literacy in Archaeology. \emph{Advances in Archaeological Practice} 10(2): 177--186. DOI: https://doi.org/\href{https://doi.org/10.1017/aap.2022.3}{10.1017/aap.2022.3}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-lambers2024}{}}%
Lambers, K. 2024. \emph{\emph{\href{https://hdl.handle.net/1887/4054691}{The many challenges of digital and computational archaeology}}}. PhD thesis.

@@

\leavevmode\vadjust pre{\hypertarget{ref-morgan2022}{}}%
Morgan, C. 2022 Current digital archaeology. \emph{Annual Review of Anthropology} 51(1): null. DOI: https://doi.org/\href{https://doi.org/10.1146/annurev-anthro-041320-114101}{10.1146/annurev-anthro-041320-114101}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-visser2016}{}}%
Visser, RM, Zijverden, WK van and Alders, PG. 2016 \emph{Teaching digital archaeology digitally}. In: Campana, S, Scopigno, R, Carpentiero, G, and Cirollo, M (eds.). Oxford: Archaeopress. pp. 11--15.
